|

### 2023-03-22 paper/01-introduction.Rmd
begin dumping in text for data section to clean up and modify later

The first BDB led to several contributions largely focused offensive performance evaluation. For example. finalists @deshpande2020expected introduced an approach for modeling the hypothetical completional probability of a pass aiding in the evalaution of quarterback (QB) decision making. The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet]. Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going]. Since its inception, the BDB has changed themes each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams. The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays.

>>>>

The first BDB led to several contributions largely focused offensive performance evaluation. For example. finalists @deshpande2020expected introduced an approach for modeling the hypothetical completional probability of a pass aiding in the evalaution of quarterback (QB) decision making. The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet]. Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going]. Since its inception, the BDB has changed themes each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams. The [2023 edition of the NFL BDB](https://www.kaggle.com/competitions/nfl-big-data-bowl-2023) continued this revolving theme by asking participants to evaluate lineman on passing plays [@Howard2023NFL].

### 2023-03-22 paper/02-data.Rmd
begin dumping in text for data section to clean up and modify later

We rely on Next Gen Stats data provided by the 2023 NFL Big Data Bowl competition [@Howard2023NFL] for our investigation.
The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.
Also included are general information about the players, games, and plays within each game
At a more granular level, we are given information on the player and football locations recorded at 10 frames per second for each play.
Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.

>>>>

We use the tracking data made available in the 2023 NFL BDB competition for our investigation. The data span the first eight weeks of the 2021 NFL regular season, consisting of a total of 122 games.

Also included are general information about the players, games, and plays within each game
At a more granular level, we are given information on the player and football locations recorded at 10 frames per second for each play.
Also included are scouting data supplied by Pro Football Focus, which collects player-level information at the play level.

In December 2018, the NFL became the first North American professional sports league to release a portion of their tracking data to the public when temporarily made available a subset of this data from the first 6 weeks of the 2017 season for the inaugural “Big Data Bowl” competition.[2]

The NFL’s tracking data collected as follows: Two radio frequency identification (RFID) chips are placed in each player’s shoulder pads (and in the ball). The RFID chips emit a signal to sensors in each stadium, which triangulate the location of the chip on the field. The data is collected at a rate of 10 Hz, so that the on-field location, speed, and angle of each player (and the ball) is recorded 10 times per second. Event annotations (e.g. ball snapped, first contact, pass thrown, etc) are recorded by the NFL for each play. In total, the dataset contains 1,075,720 unique frames across 14,167 plays, each of which records the locations and trajectories (speed, angle) of all twenty-two players (and the ball) on the field.

Table 1 shows an example of this data for a 47-yard touchdown run by WR Cordarrelle Patterson, which occurred in a Week 6 game between the Los Angeles Chargers and Oakland Raiders in the 2017 season. Four frames from this play are displayed in Figure 2 displaying the coordinates of the offense (blue), defense (orange), and the ball-carrier (black) at particular events in the play. We will visualize the player tracking data in this manner for the remainder of the manuscript.

This data can easily be joined to existing play-by-play data from the NFL’s API (e.g. via the nflscrapR package), which contains additional information about each play (Horowitz et al. 2017). For the models in Section 4, we identified all ball-carrier sequences for running plays, which includes designed runs and QB scrambles. While the tracking data records the location of the ball in addition to the players, it does not identify who is the ball-carrier for a particular frame. We first identified the ball-carriers for every type of play (pass attempts, runs, returns, etc.) based on the information available from the NFL’s API via nflscrapR, which denotes who was directly involved in each play. Given the roles a player can have (passer, runner, receiver, interceptor, or returner), we used the provided event annotations to determine when a player became the ball-carrier. Since, for simplicity, we focus our attention on running plays in this manuscript, we identify the beginning of the ball-carrier sequence when the runner received the ball by either a handoff, lateral, or direct snap. The end of the ball-carrier sequence was marked when either the player was tackled, ran out of bounds, fumbled, or scored a touchdown. We excluded all plays missing the necessary information from the NFL API, as well as plays where the snap of the play was missing in the tracking data, and any ball-carrier sequences where either the starting or ending events were missing. After further pre-processing for the covariates described in Section 4, our final modeling dataset consisted of 154,908 frames from 4502 unique ball-carrier sequences on running plays. Figure 3A displays the distribution of the length of these ball-carrier sequences, revealing that majority of ball-carrier sequences are between two to five seconds in length, while Figure 3B displays the observed change in field position from the ball-carrier’s current location that will be modeled, as discussed in 3.3.

The NFL Big Data Bowl Dataset contains tracking, play, and game data from all 91 games
in the the first 6 weeks of the 2017 NFL season. The tracking data is at the granularity of
every tenth of a second per play. For each player on the field (and the ball) for a given play,
5
the data contains: time stamp of play (time, yyyy-mm-dd, hh:mm:ss), player position along
the long axis of the field (0 - 120 yards), player position along the short axis of the field
(0 - 53.3 yards), speed in yards/second, distance traveled from prior time point (in yards),
angle of player motion (0 - 360 degrees), tagged play details, (including moment of ball snap,
pass release, pass catch, tackle, etc), player identification number (unique across players),
player name, jersey number of player, team (away or home) of corresponding player, frame
identifier for each play (starting at 1), unique game identifier, and play identifier (not unique
across games). From the tracking data we are able to calculate eadch player’s distance to
the ball and to other players as well as their cumulative distance run in the play and in the
game. The play data contains (not an exhaustive list): game quarter, time on game clock at
the start of the play (counting down from 15:00, MM:SS), down, distance needed for a first
down, yard line at line-of-scrimmage, home team score prior to the play, visiting team score
prior to the play, home team points at the end of the play, visiting team points at the end
of the play, indicator for penalty called on play, indicator for special teams play, pass length
(in yards), result of pass play (caught, incomplete, intercepted, run, sack), result of play in
yards, and a description of the play. The play data allows us to calculate the difference in
score and incorporate the timing of the play in our models. The game data contains game
specific information like final score, temperature, humidity, and wind. We did not use any
of this information in our models. More detailed information on the data can be found at
the Big Data Bowl GitHub page
